Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-03-29 pubs/tfp16/paper.tex
pm edits

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
       (New-project-in-progress Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance.
\\
\emph{Preliminatry paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the workshop.  The fault-tolerance mechanisms
    remain future work.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade,~\cite{Exascale-roadmap} a highly
respected member of the HPC community recently stated publicly that ``Fortran
is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
  Japanese, and international groups are also anticipating the realization
  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran and
complete rewrites are infeasible.~\footnote{The codes of our lab's British
  counterpart are reportedly in the same state.}  C++ has become more common
for new code starts, but their nature remains the same: huge codes with fair
to non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be poor and is explained by
Amdahl's law because the parallel model \emph{as typically implemented} is
bulk synchronous.

@@

A common complaint among these budding high-performance functional programmers
is the need to get strictness just right: to remove space leaks, achieve
acceptable performance, get communication (e.g.\ MPI) and IO to work as they
expect, and so on.  More specifically, they expect to be able to reason about
space and time usage using their existing mental models. While they are happy
to write code that is strongly reminiscent of the numerical examples in
Hughes' classic piece (whether they've read it or not)~\cite{Hughes89}, they
would also be happy to write in a slightly modified style that does not
require non-strictness---after all, that's how they've been thinking their
entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with lazy (or non-strict) by default semantics, with the most telling recent
concrete evidence being the push for, and realization of, the \texttt{Strict}
and \texttt{StrictData} language extensions newly available in GHC 8.0.
Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer scientists}.
The concept of a space leak, for example, caused by other than a missing explicit
deallocation statement (C \emph{free}, etc.) is difficult to understand and not
something they particularly want to think about.  Hence, we eschew explicit
strictness annotations for varying uniform degrees of strictness.

We claim, then, that there is a trend towards \emph{strict-by-default pure
  functional programming}, whether as imposed by programmer discipline, a parallel
runtime system (such as the aforementioned Legion, wherein the pure semantics
at function granularity allows the runtime system to relax the nominal serial
ordering), or the language itself (e.g.\ strict Haskell).

@@

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack---the scalability of checkpoint/restart may be reaching its
end.  In the imperative world side-effect-free tasks could be safely restarted
after failure, and the Legion system is being augmented with this capability.
Ericsson's quasi-functional language Erlang and its runtime system was
designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been recognized
that the pure parts of a functional program (excluding, e.g., the I/O monad
in Haskell) could be safely restarted.

@@

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict.  Light-weight has several
meanings or implications.  First and foremost it means that \emph{we are not
  attempting to compete with ghc in any respect}---succumbing to the
temptation of recreating ghc language, type system, and runtime features, or
significant code optimization, would simply divert us from our central
goals.\footnote{We estimate that we will have, to within an order of
  magnitude, one person-hour of total available development time for each
  person-year that ghc currently represents, or put another way, one second
  for each hour, respectively.}

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

A secondary, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of
non-strictness.\footnote{The starting experience of our student interns to
  date has been normalized by having them complete Hutton's \emph{Programming
    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
  substituting his article \emph{Higher-order Functions for
    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
  book.}

@@

First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense.  Similarly, when referring
to language semantics we refer to the operational rather than denotational
semantics.

@@

ghc 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor,
strictness.  In Haskell the distinction between constructors and functions is
somewhat blurred because constructors behave like functions (outside of
pattern matching), but this need not be the case.  In STG---one of ghc's
intermediate representations---the distinction is clear: heap objects are
explicitly constructed by \texttt{let} constructs in which constructor names
are given explicitly (not aliased); constructor names are not functions.

@@

Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
is part of our master plan.  In short, our plan is to use ghc as our front-end
to generate its intermediate form, STG, then escape to our system from there.
There are various possible ways to accomplish the former: using ghc's option
to dump STG as text, hacking ghc itself
and, optionally, using ghc as a library.  We note that this approach has
a precedent in the aforementioned Nvidia research compiler.

@@

The STG language is ...

Our incarnation of STG is very similar to that given in EA, as is
our presentation which is meant to evoke both an abstract and concrete syntax.

\begin{table}
\footnotesize % tiny scriptsize footnotesize small
\centering
\begin{tabular}{r r c l l}

Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
Constructor  & $C$            &     &                                              & Initial upper-case letter \\
Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point literal \\
Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
\\
Expression   & $e$            & ::= & $a$                                          & Atom \\
             &                & $|$ & $f\ a_1\dots a_n$                            & Application, $n\ge 1$ \\
             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$
                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
             &                & $|$ & \texttt{case} $e$ \texttt{as}
                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n \ge 1$\\
\\
Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
\\
Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
             &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
\\
Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $1\le i \le n \ge 1$, distinguished $\mathit{main}$ \\

\end{tabular}
\end{table}

While our interpretation of STG is as a simple higher-order, polymorphic, pure
functional language, some syntactic restrictions make direct programming
rather tedious.  This is discussed further in a later section.

Operationally, when an application such as

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG machine described later.  Our description is based on that given in
Marlow and Peyton Jones' \emph{fast curry} paper~\cite{Marlow:2004}, to which we
refer so many times that we will subsequenty refer to it as
\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely
  has another connotation to the reader.}

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
regardless of what argument type it might be dynamically applied to.  We
distinguish \emph{application}, e.g.\ \texttt{f x y}, which is a syntactic
notion, from actually calling a function. Thus \texttt{id id x} is an
application of \texttt{id} to two arguments, but the underlying \emph{FUN}
(user- or system- defined function) when actually called is given exactly one
argument.

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments already provided.  If the arity is greater than two then a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two then the underlying
\emph{FUN} is tail-called with any previous argument (in case of \emph{PAP})
and new arguments \texttt{x} and \texttt{y}.  If the arity is less than two
(it must be at least one, so one in this case) the underlying \emph{FUN} is
called-with-return with any previous argument (in case of \emph{PAP}) and new
argument \texttt{x}, then the object returned applied to remaining argument
\texttt{y}.

@@

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict, fully strict---all arguments
in an application are evaluated---and the intermediate in which only those
arguments needed by a function call are evaluated.  The STG machinery makes
the choice of evaluation strategy easy to change, both in terms of function
and constructor application.

@@

\subsubsection{Proper tail calls}

While performance is not a high priority, we attain proper tail calling instead of
using the trampoline used in earlier versions of ghc~\cite{PJ-stockhardware}.
Because we maintain our own data and control (continuation) stack all
generated C functions take no explicit arguments and have return type void,
and with a single exception in the non-strict case are invoked in tail call
positions, at non-zero optimization levels both gcc and Clang/LLVM generate
jumps instead of calls for C calls in tail call positions.  When using gcc
we also use a fixed set of machine registers for passing/returning values.

The exceptions to tail calling are for convenience.  Similar to what is
described in the \emph{EA} we have a generic variadic stgApply function that
is used for applications of non-known functions, and known functions with a
deficit or excess number of arguments.  The stgApply function takes as
arguments the first element of the application (FUN, PAP, THUNK, or BLACKHOLE)
and its arguments, and if a THUNK, evaluates it with a call (rather than jump)
to yield a FUN or PAP.  This allows us to implement a single stgApply function
as a single C function taking a contiguous block of machine words (unboxed
values or pointers) on the explicitly stack.  The implementation of stgApply
as described in \emph{EA} makes the same concession regarding
call-with-return, but the scheme described there uses an auto-generated family
of stgApply functions parameterized by number and type (unboxed or pointer) of
arguments.  We used this scheme initially as well, but because the parameter
block contains a bitmap describing which arguments are boxed/unboxed for the
benefit of the garbage collector, this elaboration is unnecessary.

For stgApply in the non-strict semantics an alternative to call-with-return
would be a pair of functions stgApply1 and stgApply2.  In one possible variation,
stgApply1 would push a continuation with stgApply2 as the return address, then
invoke the first element of the application.  Call-with-return becomes more
convenient when considering the various forms of strictness:  when evaluating
function or constructor arguments it is simpler, and arguably more efficient
in terms of time (but not space in general) to ``evaluate in place'' than
implement the continuation style to the logical limit.

\subsection{Mini-Haskell}

As mentioned, a current near-term goal is to use ghc as our front-end
to generate STG as input to our system.  In the meantime, as a stop-gap
measure to make the generation of test cases easier, we have implemented
a \emph{mini-Haskell} front-end, a strict subset of standard Haskell syntax.

First while the scrutinee of a case expression may be an arbitrary expression,
arguments in function applications, constructor allocation, and primitive
operations are constrained to be atoms.  For unboxed types a case
a trivial case expression to alleviates
the difficulty.
\begin{verbatim}
  f ... e ... -> case e as x of _ -> f ... x ...
\end{verbatim}

{\color{red} I didn't want to touch the above text, but there's obviously some 
redundancy in the content here...}

Writing STG in the concrete syntax conceived for our project is tedious and
prone to errors.  Compared to Haskell, the eventual source language for the
compiler, STG is relatively verbose, requiring explicit block and expression
delimiters, and classification of heap objects. Though its general structure is
similar to primitive Haskell, it lacks many of the conveniences that make
Haskell a pleasure to work with.  The motivation for a more expressive language
for testing our compiler in its nascent stages is then clear.  This language,
mini-Haskell, was progressively "grown" from the STG syntax towards a subset of
Haskell itself*, removing much of the frustrating syntactic cruft and adding
some of the core features that are often taken for granted in Haskell (and other
languages).

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
       (New project paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the workshop.  The fault-tolerance mechanisms
    remain future work.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade~\cite{Exascale-roadmap} a highly
respected member of the HPC community recently stated publicly that ``Fortran
is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
  Japanese, and international groups are also anticipating the realization
  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are reportedly in the same state.}  C++ has become more common
for new code starts, but their nature remains the same: huge codes with fair
to non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be poor and is explained by
Amdahl's law because the parallel model \emph{as typically implemented} is
bulk synchronous.

@@

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of execution, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece~\cite{Hughes89}
(whether they've read it or not), they would also be happy to write in a
slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with lazy (or non-strict) by default semantics, with the most telling recent
concrete evidence being the push for, and realization of, the \texttt{Strict}
and \texttt{StrictData} language extensions newly available in GHC 8.0.
Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer scientists}.
The concept of a space leak, for example, caused by other than a missing explicit
deallocation statement (C \emph{free}, etc.) is difficult to understand and not
something they particularly want to try to understand.  Hence, we eschew explicit
strictness annotations for varying uniform degrees of strictness.

@@

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various point in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can restart from the latest stored
state.  Most implementations are explicit, though there has been some success
with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the programmer.
Ericsson's quasi-functional language Erlang and its runtime system was
designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been
recognized that the pure parts of a functional program (excluding, e.g., the
I/O monad in Haskell) could be safely restarted.  The recent work of
[Trinder's student] impressively demonstrates this for an augmented ghc/runtime
on a large-scale distributed-memory system.

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming}, whether as imposed by
programmer discipline, a parallel runtime system, or the language itself
(e.g.\ strict Haskell).  Second, we claim that there is a trend in
high-performance computing, and by transitivity functional programming,
towards transparent fault tolerance.

@@

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict, and secondarily, with mechanisms
for transparent fault tolerance.  Light-weight has several meanings or
implications.  First and foremost it means that \emph{we are not attempting to
  compete with ghc in any respect}---succumbing to the temptation of
recreating ghc language, type system, and runtime features, or significant
code optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that ghc currently
  represents, or put another way, one second for each hour, respectively.}

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\end{itemize}

@@

A minor, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of
non-strictness.\footnote{The starting experience of our student interns to
  date has been normalized by having them complete Hutton's \emph{Programming
    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
  substituting his article \emph{Higher-order Functions for
    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
  book.}

@@

First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

ghc 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor,
strictness.  In Haskell the distinction between constructors and functions is
somewhat blurred because constructors behave like functions (outside of
pattern matching), but this need not be the case.

@@

Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
is part of our master plan.  In short, our plan is to use ghc as our front-end
to generate its intermediate form, STG~\cite{PJ-stockhardware}, then escape to
our system from there.  There are various possible ways to accomplish the
former: using ghc's option to dump STG as text, hacking ghc itself and,
optionally, using ghc as a library.

@@

Our incarnation of STG, as is our presentation in Figure~\ref{STGsyntax} which
is meant to evoke both an abstract and concrete syntax, are very similar to
that given in Marlow and Peyton Jones' \emph{fast curry}
paper~\cite{Marlow:2004}, to which we refer so many times that we will
subsequently refer to it as simply \emph{EA}.\footnote{\emph{Eval/Apply} from
  the title because \emph{FC} likely has another connotation to the reader.}

\begin{figure}
%\centering
\begin{centering}
\footnotesize % tiny scriptsize footnotesize small
\begin{tabular}{r r c l l}

Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
Constructor  & $C$            &     &                                              & Initial upper-case letter \\
Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point literal \\
Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
\\
Expression   & $e$            & ::= & $a$                                          & Atom \\
             &                & $|$ & $f\ a_1\dots a_n$                            & Application, $n\ge 1$ \\
             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\

             &                & $|$ & \texttt{let \{}                              & Recursive let, $n\ge 1$ \\
             &                &     & \texttt{ } $x_1$ \texttt{=} $\mathit{obj}_1$ \texttt{;} \\
             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
             &                &     & \texttt{ } $x_n$ \texttt{=} $\mathit{obj}_n$ \\
             &                &     & \texttt{\} in} $e$  \\

             &                & $|$ & \texttt{case} $e$ \texttt{of} $x$ \texttt{\{}  & Case expression, $n \ge 1$\\
             &                &     & \texttt{ } $\mathit{alt}_1$ \texttt{;} \\
             &                &     & \hspace{0.2in} $\dots$ \texttt{;} \\
             &                &     & \texttt{ } $\mathit{alt}_n$ \\
             &                &     & \texttt{\}} \\
\\
Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
\\
Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
             &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
\\
Program      & $\mathit{prog}$& ::= & $f_1\ =\ \mathit{obj}_1$ \texttt{;}          & $n \ge 1$, distinguished \texttt{main}\\
             &                &     & \texttt{ } $\dots$ \texttt{;} \\
             &                &     & $f_n\ =\ \mathit{obj}_n$

\end{tabular}
\end{centering}
\caption{STG syntax}
\label{STGsyntax}
\end{figure}

Partial application (\emph{PAP}) and black hole (\emph{BLACKHOLE}), i.e., a thunk being
evaluated, are only runtime entities and not actually part of our STG language.

The semantics of STG is what one would expect for a higher-order, polymorphic,
pure functional language, with one exception for a Haskell programmer familiar
with unboxed types:  \texttt{case} is strict in the scrutinee, so
\texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in Haskell,
that is, the scrutinee is evaluated first.

In STG---one of ghc's
intermediate representations---the distinction is clear: heap objects are
explicitly constructed by \texttt{let} constructs in which constructor names
are given explicitly (not aliased); constructor names are not functions.

A distinctive feature of STG is that heap allocation is explicit; this is exactly
what \texttt{let} bindings do.  As such unboxed values cannot be let-bound; instead
unboxed values can be bound with \texttt{case}:
\begin{verbatim}
    case e of x# {
      _ -> e
    }
\end{verbatim}
where the \texttt{\#} suffix is merely a convention to signify an unboxed type.
Deferred evaluation is made explicit:  a deferred expression evaluation is
encoded as a thunk.

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG.  Our partial description is based on that given in \emph{EA}.

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
regardless of the type of the argument (e.g., some function type) it might be
applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
a syntactic notion, from actually calling a function. Thus \texttt{id id x} is
an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
(user- or system- defined function) when actually called is given exactly one
argument (the first one).

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments previously provided.  If the arity is greater than two then a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two then the underlying
\emph{FUN} is tail-called with any previous argument (in case of \emph{PAP})
and new arguments \texttt{x} and \texttt{y}.  If the arity is less than two
(it must be at least one, so one in this case) the underlying \emph{FUN} is
called-with-return with any previous arguments (in case of \emph{PAP}) and new
argument \texttt{x}, then the object returned applied to remaining argument
\texttt{y}.

@@

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict, fully strict---all arguments
in an application are evaluated---and the intermediate in which only those
arguments needed by a function call are evaluated.  The STG machinery makes
the choice of evaluation strategy easy to change, both in terms of function
application and building a constructor in the heap.

@@

\subsubsection{Proper tail calls.}
While performance is not a high priority, we attain proper tail calling instead of
using the trampoline used in earlier versions of ghc~\cite{PJ-stockhardware}.
Because we maintain our own data and control (continuation) stack all
generated C functions take no explicit arguments and have return type void,
and with a single exception in the non-strict case are invoked in tail call
positions, at non-zero optimization levels both gcc and Clang/LLVM generate
jumps instead of calls for C calls in tail call positions.  When using gcc
we also use a fixed set of machine registers for passing/returning values.

The exceptions to tail calling are for convenience.  Similar to what is
described in \emph{EA} we have a generic variadic \emph{stgApply} function
that is used for applications of non-known functions, and known functions with
a deficit or excess number of arguments.  The \emph{stgApply} function takes
as arguments the first element of the application (\emph{FUN}, \emph{PAP},
\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
evaluates it with a call (rather than jump) to yield a \emph{FUN} or
\emph{PAP}.  This allows us to implement a single \emph{stgApply} function as
a C function taking a contiguous block of machine words (unboxed values
or pointers) on the explicit stack.  The implementation of \emph{stgApply}
as described in \emph{EA} makes the same concession regarding
call-with-return, but the scheme described there uses an auto-generated family
of \emph{stgApply} functions parameterized by number and type (unboxed or
pointer) of arguments.  We used this scheme initially as well, but because the
parameter block contains a bitmap describing which arguments are boxed/unboxed
for the benefit of the garbage collector, this elaboration is unnecessary.

For \emph{stgApply} in the non-strict semantics an alternative to
call-with-return would be a pair of functions \emph{stgApply1} and
\emph{stgApply2}.  In one possible variation, \emph{stgApply1} would push a
continuation with \emph{stgApply2} as the return address, then invoke the
first element of the application.  Call-with-return becomes more convenient
when considering the various forms of strictness: when evaluating function or
constructor arguments it is simpler, and arguably more efficient in terms of
time (but not space in general) to ``evaluate in place'' than implement the
continuation style to the logical limit.

\subsection{Mini-Haskell}
As mentioned, a current near-term goal is to use ghc as our front-end to
generate STG as input to our system.  In the meantime, as a stop-gap measure
to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote an
unboxed literal value.  Our extension (to STG as well)
is the provision of the optional \texttt{unboxed} keyword in datatype definitions
to allow user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}
